### 2025-02-11 paper/preprint/paper.tex
just experimental setup and stuff left now and then sending off the first draft

\usepackage{arxiv}
\usepackage{orcidlink}
\usepackage{amsmath}
\usepackage[T1]{fontenc}
\usepackage{placeins}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\usepackage{amsthm}
\theoremstyle{definition}
\newtheorem{exercise}{Parameters}[section]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\theoremstyle{remark}
\AtBeginDocument{\renewcommand*{\proofname}{Proof}}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\newtheorem{refremark}{Remark}[section]
\newtheorem{refsolution}{Solution}[section]
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

@@

Since actionability constraints are binding at test time, we should also
impose them when generating \(\mathbf{x}^\prime\) during each training
iteration to align model representations with user requirements. Through
their effect on \(\mathbf{x}^\prime\), both types of constraints
influence model outcomes through Equation~\ref{eq-div}. This is best
explained through a simple example involving a linear classifier:
\(\mathbf{M}_\theta(\mathbf{x})=\theta^\intercal\mathbf{x}\). Then for
\(y^+ \sim p(y)\) we have:

\begin{equation}\phantomsection\label{eq-grad}{
\nabla_\theta \left(\text{div}(\mathbf{x},\mathbf{x^\prime},\mathbf{y};\theta)\right) = \nabla_\theta \left(-\mathbf{M}_\theta(\mathbf{x})[y^+]\right) - \nabla_\theta \left(-\mathbf{M}_\theta(\mathbf{x}^\prime)[y^+]\right) = \mathbf{x}^\prime - \mathbf{x}
}\end{equation}

In other words, contrastive divergence will push down (up) on the
coefficient \(\theta[d]\) corresponding to a feature \(d\) that is lower
(higher) for counterfactuals than samples. This has the intended effect
of aligning counterfactual explanations with the data, as discussed
above.

Crucially though, we want to avoid penalizing implausibility that arises
due to mutability constraints. Suppose, for example, that feature \(d\)
represents \emph{age} and our model predicts the likelihood of credit
default.

@@

@@

>>>>

\usepackage{arxiv}
\usepackage{orcidlink}
\usepackage{amsmath}
\usepackage[T1]{fontenc}
\usepackage{placeins}
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\@ifpackageloaded{fontawesome5}{}{\usepackage{fontawesome5}}
\definecolor{quarto-callout-color}{HTML}{909090}
\definecolor{quarto-callout-note-color}{HTML}{0758E5}
\definecolor{quarto-callout-important-color}{HTML}{CC1914}
\definecolor{quarto-callout-warning-color}{HTML}{EB9113}
\definecolor{quarto-callout-tip-color}{HTML}{00A047}
\definecolor{quarto-callout-caution-color}{HTML}{FC5300}
\definecolor{quarto-callout-color-frame}{HTML}{acacac}
\definecolor{quarto-callout-note-color-frame}{HTML}{4582ec}
\definecolor{quarto-callout-important-color-frame}{HTML}{d9534f}
\definecolor{quarto-callout-warning-color-frame}{HTML}{f0ad4e}
\definecolor{quarto-callout-tip-color-frame}{HTML}{02b875}
\definecolor{quarto-callout-caution-color-frame}{HTML}{fd7e14}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\usepackage{amsthm}
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\theoremstyle{definition}
\newtheorem{exercise}{Parameters}[section]
\theoremstyle{definition}
\newtheorem{example}{Example}[section]
\theoremstyle{remark}
\AtBeginDocument{\renewcommand*{\proofname}{Proof}}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\newtheorem{refremark}{Remark}[section]
\newtheorem{refsolution}{Solution}[section]
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

@@

Since actionability constraints are binding at test time, we should also
impose them when generating \(\mathbf{x}^\prime\) during each training
iteration to align model representations with user requirements. Through
their effect on \(\mathbf{x}^\prime\), both types of constraints
influence model outcomes through Equation~\ref{eq-div}. Here it is
crucial that we avoid penalizing implausibility that arises due to
mutability constraints. For any mutability-constrained feature \(d\)
this can be achieved by enforcing
\(\mathbf{x}[d] - \mathbf{x}^\prime[d]:=0\) whenever perturbing
\(\mathbf{x}^\prime[d]\) in the direction of \(\mathbf{x}[d]\) would
violate mutability constraints. Specifically, we set
\(\mathbf{x}[d] := \mathbf{x}^\prime[d]\) if

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Feature \(d\) is strictly immutable in practice.
\item
  We have \(\mathbf{x}[d]>\mathbf{x}^\prime[d]\) but feature \(d\) can
  only be decreased in practice.
\item
  We have \(\mathbf{x}[d]<\mathbf{x}^\prime[d]\) but feature \(d\) can
  only be increased in practice.
\end{enumerate}

From a Bayesian perspective, setting
\(\mathbf{x}[d] := \mathbf{x}^\prime[d]\) can be understood as assuming
a point mass prior for \(p(\mathbf{x})\) with respect to feature \(d\).
Intuitively, we think of this simply in terms ignoring implausibility
costs with respect to immutable features, which effectively forces the
model to instead seek plausibility with respect to the remaining
features. This in turn results in lower overall sensitivity to immutable
features, which we demonstrate empirically below. Under certain
assumptions this can also be shown to hold analytically
Section~\ref{sec-app-constraints}.

@@

\bibitem[\citeproctext]{ref-hastie2009elements}
Hastie, Trevor, Robert Tibshirani, and Jerome Friedman. 2009. \emph{The
Elements of Statistical Learning}. Springer New York.
\url{https://doi.org/10.1007/978-0-387-84858-7}.

@@

\subsection{Details on Encoding Mutability
Constraints}\label{sec-app-constraints}

In Section~\ref{sec-constraints} we explain that to avoid penalizing
implausibility that arises due to mutability constraints, we impose a
point mass prior on \(p(\mathbf{x})\) for the corresponding feature. We
argue in Section~\ref{sec-constraints} that this approach induces models
to be less sensitive to feature \(d\) vis-à-vis some other mutable
feature \(m\) and demonstrate this empirically in
Section~\ref{sec-experiments}. Below we derive this result analytically
for linear classifier.

\subsubsection{Linear Classifier}\label{linear-classifier}

Let \(\mathbf{M}_\theta(\mathbf{x})=\Theta\mathbf{x}\) denote a linear
classifier with \(y\in\{1,...,K\}=\mathcal{K}\) and
\(\mathbf{x} \in \mathbb{R}^D\). Let \(d_{\text{mtbl}}\) and
\(d_{\text{immtbl}}\) denote some mutable and immutable feature,
respectively. Suppose that
\(\mu_{y^-,d_{\text{immtbl}}} < \mu_{y^+,d_{\text{immtbl}}}\) and
\(\mu_{y^-,d_{\text{mtbl}}} > \mu_{y^+,d_{\text{mtbl}}}\), where
\(\mu_{k,d}\) denotes the conditional sample mean of feature \(d\) in
class \(k\).

Assuming multivariate Gaussian class densities with common diagonal
covariance matrix \(\Sigma_k=\Sigma\) for all \(k \in \mathcal{K}\), we
have for the log likelihood ratio between any two classes
\(k,m \in \mathcal{K}\) (\citeproc{ref-hastie2009elements}{Hastie,
Tibshirani, and Friedman 2009}):

\begin{equation}\phantomsection\label{eq-loglike}{
\log \frac{p(k|x_{d})}{p(m|x_{d})}=\frac{\mu_{k,d}-\mu_{m,d}}{\sigma_{d}^2} x_{d} + \text{const}
}\end{equation}

By Equation~\ref{eq-loglike} we have the
\(\theta_{y^-,d_{\text{immtbl}}} - \theta_{y^+,d_{\text{immtbl}}} < 0\)
and \(\theta_{y^-,d_{\text{mtbl}}} - \theta_{y^+,d_{\text{mtbl}}} > 0\),
where \(\theta_{k,d}=\Theta[k,d]\) denotes the coefficient on feature
\(d\) for class \(k\).

Let \(\mathbf{x}^\prime\) denote some randomly chosen individual from
class \(y^-\) and let \(y^+ \sim p(y)\) denote the randomly chosen
target class. Then the partial derivative of the contrastive divergence
penalty Equation~\ref{eq-div} with respect to coefficient
\(\theta_{y^+,d}\) is equal to

\begin{equation}\phantomsection\label{eq-grad}{
\frac{\partial}{\partial\theta_{y^+,d}} \left(\text{div}(\mathbf{x},\mathbf{x^\prime},\mathbf{y};\theta)\right) = \frac{\partial}{\partial\theta_{y^+,d}} \left( \left(-\mathbf{M}_\theta(\mathbf{x})[y^+]\right) - \left(-\mathbf{M}_\theta(\mathbf{x}^\prime)[y^+]\right) \right) = x_{d}^\prime - x_{d}
}\end{equation}

and equal to zero everywhere else.

Since \(\mu_{y^-,d_{\text{immtbl}}} < \mu_{y^+,d_{\text{immtbl}}}\) we
are more likely to have
\(x_{d_{\text{immtbl}}}^\prime - x_{d_{\text{immtbl}}} < 0\) than vice
versa at initialization. Similarly, we are more likely to have
\(x_{d_{\text{mtbl}}}^\prime - x_{d_{\text{mtbl}}} > 0\) by
\(\mu_{y^-,d_{\text{mtbl}}} > \mu_{y^+,d_{\text{mtbl}}}\).

This implies that if we do not protect feature \(d_{\text{immtbl}}\),
the contrastive divergence penalty will push down on
\(\theta_{y^-,d_{\text{immtbl}}}\) thereby exacerbating the existing
effect
\(\theta_{y^-,d_{\text{immtbl}}} - \theta_{y^+,d_{\text{immtbl}}} < 0\).
In words, not protecting the immutable feature would have the
undesirable effect of making the classifier more sensitive to this
feature, in that it would be more likely to predict class \(y^-\) as
opposed to \(y^+\) for lower values of \(d_{\text{immtbl}}\).

By the same rationale, the contrastive divergence penalty can generally
be expected to push up on \(\theta_{y^-,d_{\text{mtbl}}}\) exacerbating
\(\theta_{y^-,d_{\text{mtbl}}} - \theta_{y^+,d_{\text{mtbl}}} > 0\). In
words, this has the effect of making the classifier more sensitive to
the mutable feature, in that it would be more likely to predict class
\(y^-\) as opposed to \(y^+\) for higher values of \(d_{\text{mtbl}}\).

Thus, our proposed approach of protecting feature \(d_{\text{immtbl}}\)
has the net affect of decreasing the classifier's sensitivity to the
immutable feature relative to the mutable feature (i.e.~no change in
sensitivity for \(d_{\text{immtbl}}\) relative to increased sensitivity
for \(d_{\text{mtbl}}\)).

We provide an illustrative example in Example~\ref{exm-grad}.

\begin{example}[Prediction of Consumer Credit
Default]\protect\hypertarget{exm-grad}{}\label{exm-grad}

Suppose now that \(d_{\text{immtbl}}\) represents an individual's
\emph{age} and \(d_{\text{mtbl}}\) represents an individual's existing
level of credit card debt. Assume that these two features are
independent, the class conditional densities are Gaussian and we use no
other features to predict the risk of individuals defaulting on a
consumer loan using a linear classifier. We have simulated this scenario
using synthetic data in Figure~\ref{fig-mtblty}.

In panel (a) of Figure~\ref{fig-mtblty}, we have trained the linear
classifier using counterfactual training, treating both features as
mutable. The linear decision boundary is roughly equally sensitive to
both \emph{age} and existing levels of \emph{debt}.

Conversely, in panel (b) of Figure~\ref{fig-mtblty}, we have trained the
same classifier using counterfactual training, but this time treating
\emph{age} as an immutable feature. The result is a new decision
boundary that has tilted in favor of higher sensitivity to the mutable
feature (existing \emph{debt}) and lower sensitivity to the immutable
feature (\emph{age}).

\end{example}

\begin{figure}

\centering{

\pandocbounded{\includegraphics[keepaspectratio]{paper_files/mediabag/../../paper/figures/app_mtblty.pdf}}

}

\caption{\label{fig-mtblty}Visual illustration of the effect of imposing
mutability constraints. See Example~\ref{exm-grad} for details.}

\end{figure}%

\begin{tcolorbox}[enhanced jigsaw, coltitle=black, opacityback=0, rightrule=.15mm, leftrule=.75mm, colback=white, left=2mm, colframe=quarto-callout-warning-color-frame, opacitybacktitle=0.6, breakable, toprule=.15mm, colbacktitle=quarto-callout-warning-color!10!white, bottomrule=.15mm, toptitle=1mm, titlerule=0mm, arc=.35mm, title=\textcolor{quarto-callout-warning-color}{\faExclamationTriangle}\hspace{0.5em}{Warning}, bottomtitle=1mm]

@Cynthia, @Arie, let me know if you want to phrase the above in terms of
a theorem and proof. This is something I've so far shied away from, but
I can see how it might be appropriate here.

\end{tcolorbox}

\section{Detailed Results}\label{detailed-results}

### 2025-02-11 paper/sections/method.qmd
just experimental setup and stuff left now and then sending off the first draft

From a Bayesian perspective, setting $\mathbf{x}[d] := \mathbf{x}^\prime[d]$ can be understood as assuming a point mass prior for $p(\mathbf{x})$ with respect to feature $d$. Intuitively, we think of this simply in terms ignoring implausibility costs with respect to immutable features, which effectively forces the model to instead seek plausibility with respect to the remaining features. This in turn results in lower overall sensitivity to immutable features, which we demonstrate empirically bellow and analytically for the linear case in @sec-app-constraints of the supplementary appendix.

>>>>

From a Bayesian perspective, setting $\mathbf{x}[d] := \mathbf{x}^\prime[d]$ can be understood as assuming a point mass prior for $p(\mathbf{x})$ with respect to feature $d$. Intuitively, we think of this simply in terms ignoring implausibility costs with respect to immutable features, which effectively forces the model to instead seek plausibility with respect to the remaining features. This in turn results in lower overall sensitivity to immutable features, which we demonstrate empirically below. Under certain assumptions this can also be shown to hold analytically [@sec-app-constraints].

### 2025-02-11 paper/sections/other/constraints.qmd
just experimental setup and stuff left now and then sending off the first draft

```{julia}
projectdir = splitpath(pwd()) |>
    ss -> joinpath(ss[1:findall([s == "CounterfactualTraining.jl" for s in ss])[1]]...) 
cd(projectdir)
```

```{julia}
using CTExperiments
using CTExperiments.CounterfactualExplanations
using CTExperiments.CounterfactualTraining
using CTExperiments.Flux
using CTExperiments.Plots
using CTExperiments.TaijaPlotting
```

# Linear Model

First, we define the mutability constraints and load the data and model:

```{julia}
using CTExperiments: 
    get_ce_data, 
    train_val_split, 
    build_model, 
    LinearModel, 
    get_input_encoder

# Data:
constraints = ["none", "both"]        # x1 cannot be mutated
# constraints = ["both", "none"]        # x2 cannot be mutated
data = LinearlySeparable(
    n_train=500,
    batchsize=50,
    mutability=constraints
)
ce_data = get_ce_data(data)
val_size = data.n_validation / (data.n_validation + data.n_train)
train_set, val_set, _ = train_val_split(data, ce_data, val_size)

# Model:
nin = size(first(train_set)[1], 1)
nout = size(first(train_set)[2], 1)
model = build_model(LinearModel(), nin, nout)
```

Next, we defined the training objective and the counterfactual generator. Finally, we initialize the optimization state.

```{julia}
using CounterfactualExplanations.Convergence
obj = EnergyDifferentialObjective(lambda=[1.0,1.0,0.0])
# obj = VanillaObjective()
generator = GenericGenerator()
opt_state = Flux.setup(Descent(), model)
conv = MaxIterConvergence()
```

```{julia}
using CTExperiments: get_log_reg_params, get_decision_boundary

function plot_db(model, x)
    coeff = get_log_reg_params(model)
    db = get_decision_boundary(coeff)
    plt = Plots.scatter(ce_data)
    yhat = [argmax(y) for y in eachcol(model(x))]
    Plots.scatter!(x[1,:], x[2,:], label="Counterfactuals", ms=15, shape=:star, color=yhat)
    Plots.abline!(plt, db.slope, db.intercept; lw=5, label="Dec. Boundary")
    display(plt)
end
```

Then we train the model:

```{julia}
using Random
Random.seed!(42)

model, logs = counterfactual_training(
    obj,
    model,
    generator,
    train_set,
    opt_state;
    val_set = val_set,
    nepochs = 50,
    mutability = Symbol.(constraints),
    callback = plot_db,
    nce=50,
    convergence=conv, 
)
```

Finally, we visualize the results:

```{julia}
coeff = get_log_reg_params(model)
display(coeff)
db = get_decision_boundary(coeff)
Plots.scatter(ce_data)
Plots.abline!(db.slope, db.intercept; lw=5, label="Dec. Boundary")
```

>>>>

```{julia}
#| eval: false

projectdir = splitpath(pwd()) |>
    ss -> joinpath(ss[1:findall([s == "CounterfactualTraining.jl" for s in ss])[1]]...) 
cd(projectdir)

using CTExperiments
using CTExperiments.CounterfactualExplanations
using CTExperiments.CounterfactualTraining
using CTExperiments.Flux
using CTExperiments.Plots
using CTExperiments.TaijaPlotting
```

## Details on Encoding Mutability Constraints {#sec-app-constraints}

In @sec-constraints we explain that to avoid penalizing implausibility that arises due to mutability constraints, we impose a point mass prior on $p(\mathbf{x})$ for the corresponding feature. We argue in @sec-constraints that this approach induces models to be less sensitive to feature $d$ vis-à-vis some other mutable feature $m$ and demonstrate this empirically in @sec-experiments. Below we derive this result analytically for linear classifier.

### Linear Classifier

Let $\mathbf{M}_\theta(\mathbf{x})=\Theta\mathbf{x}$ denote a linear classifier with $y\in\{1,...,K\}=\mathcal{K}$ and $\mathbf{x} \in \mathbb{R}^D$. Let $d_{\text{mtbl}}$ and $d_{\text{immtbl}}$ denote some mutable and immutable feature, respectively. Suppose that $\mu_{y^-,d_{\text{immtbl}}} < \mu_{y^+,d_{\text{immtbl}}}$ and $\mu_{y^-,d_{\text{mtbl}}} > \mu_{y^+,d_{\text{mtbl}}}$, where $\mu_{k,d}$ denotes the conditional sample mean of feature $d$ in class $k$.

Assuming multivariate Gaussian class densities with common diagonal covariance matrix $\Sigma_k=\Sigma$ for all $k \in \mathcal{K}$, we have for the log likelihood ratio between any two classes $k,m \in \mathcal{K}$ [@hastie2009elements]:

$$
\log \frac{p(k|x_{d})}{p(m|x_{d})}=\frac{\mu_{k,d}-\mu_{m,d}}{\sigma_{d}^2} x_{d} + \text{const}
$$ {#eq-loglike}

By @eq-loglike we have the $\theta_{y^-,d_{\text{immtbl}}} - \theta_{y^+,d_{\text{immtbl}}} < 0$ and $\theta_{y^-,d_{\text{mtbl}}} - \theta_{y^+,d_{\text{mtbl}}} > 0$, where $\theta_{k,d}=\Theta[k,d]$ denotes the coefficient on feature $d$ for class $k$.

Let $\mathbf{x}^\prime$ denote some randomly chosen individual from class $y^-$ and let $y^+ \sim p(y)$ denote the randomly chosen target class. Then the partial derivative of the contrastive divergence penalty [@eq-div] with respect to coefficient $\theta_{y^+,d}$ is equal to

$$
\frac{\partial}{\partial\theta_{y^+,d}} \left(\text{div}(\mathbf{x},\mathbf{x^\prime},\mathbf{y};\theta)\right) = \frac{\partial}{\partial\theta_{y^+,d}} \left( \left(-\mathbf{M}_\theta(\mathbf{x})[y^+]\right) - \left(-\mathbf{M}_\theta(\mathbf{x}^\prime)[y^+]\right) \right) = x_{d}^\prime - x_{d}
$$ {#eq-grad}

and equal to zero everywhere else.

Since $\mu_{y^-,d_{\text{immtbl}}} < \mu_{y^+,d_{\text{immtbl}}}$ we are more likely to have $x_{d_{\text{immtbl}}}^\prime - x_{d_{\text{immtbl}}} < 0$ than vice versa at initialization. Similarly, we are more likely to have $x_{d_{\text{mtbl}}}^\prime - x_{d_{\text{mtbl}}} > 0$ by $\mu_{y^-,d_{\text{mtbl}}} > \mu_{y^+,d_{\text{mtbl}}}$.

This implies that if we do not protect feature $d_{\text{immtbl}}$, the contrastive divergence penalty will push down on $\theta_{y^-,d_{\text{immtbl}}}$ thereby exacerbating the existing effect $\theta_{y^-,d_{\text{immtbl}}} - \theta_{y^+,d_{\text{immtbl}}} < 0$. In words, not protecting the immutable feature would have the undesirable effect of making the classifier more sensitive to this feature, in that it would be more likely to predict class $y^-$ as opposed to $y^+$ for lower values of $d_{\text{immtbl}}$.

By the same rationale, the contrastive divergence penalty can generally be expected to push up on $\theta_{y^-,d_{\text{mtbl}}}$ exacerbating $\theta_{y^-,d_{\text{mtbl}}} - \theta_{y^+,d_{\text{mtbl}}} > 0$. In words, this has the effect of making the classifier more sensitive to the mutable feature, in that it would be more likely to predict class $y^-$ as opposed to $y^+$ for higher values of $d_{\text{mtbl}}$.

Thus, our proposed approach of protecting feature $d_{\text{immtbl}}$ has the net affect of decreasing the classifier's sensitivity to the immutable feature relative to the mutable feature (i.e. no change in sensitivity for $d_{\text{immtbl}}$ relative to increased sensitivity for $d_{\text{mtbl}}$).

We provide an illustrative example in @exm-grad.

::: {#exm-grad}

## Prediction of Consumer Credit Default

Suppose now that $d_{\text{immtbl}}$ represents an individual's *age* and $d_{\text{mtbl}}$ represents an individual's existing level of credit card debt. Assume that these two features are independent, the class conditional densities are Gaussian and we use no other features to predict the risk of individuals defaulting on a consumer loan using a linear classifier. We have simulated this scenario using synthetic data in @fig-mtblty.

In panel (a) of @fig-mtblty, we have trained the linear classifier using counterfactual training, treating both features as mutable. The linear decision boundary is roughly equally sensitive to both *age* and existing levels of *debt*.

Conversely, in panel (b) of @fig-mtblty, we have trained the same classifier using counterfactual training, but this time treating *age* as an immutable feature. The result is a new decision boundary that has tilted in favor of higher sensitivity to the mutable feature (existing *debt*) and lower sensitivity to the immutable feature (*age*).

:::

![Visual illustration of the effect of imposing mutability constraints. See @exm-grad for details.](/paper/figures/app_mtblty.svg){#fig-mtblty}

::: {.callout-warning}

\@Cynthia, \@Arie, let me know if you want to phrase the above in terms of a theorem and proof. This is something I've so far shied away from, but I can see how it might be appropriate here.

:::

```{julia}
#| eval: false

using CTExperiments: 
    get_ce_data, 
    train_val_split, 
    build_model, 
    LinearModel, 
    get_input_encoder

# Callback:
using CTExperiments: get_log_reg_params, get_decision_boundary

function plot_db(model, x)
    coeff = get_log_reg_params(model)
    db = get_decision_boundary(coeff)
    plt = Plots.scatter(ce_data)
    yhat = [argmax(y) for y in eachcol(model(x))]
    Plots.scatter!(x[1,:], x[2,:], label="Counterfactuals", ms=15, shape=:star, color=yhat)
    Plots.abline!(plt, db.slope, db.intercept; lw=5, label="Dec. Boundary")
    display(plt)
end

# Data:
constraints = [
    ["both", "both"],
    ["both", "none"]
]
```

```{julia}
#| eval: false

using CounterfactualExplanations.Convergence
using Random
Random.seed!(42)

models = []

for constraint in constraints
    # Data:
    data = LinearlySeparable(
        n_train=500,
        batchsize=50,
        mutability=constraint
    )
    ce_data = get_ce_data(data)
    val_size = data.n_validation / (data.n_validation + data.n_train)
    train_set, val_set, _ = train_val_split(data, ce_data, val_size)

    # Model:
    nin = size(first(train_set)[1], 1)
    nout = size(first(train_set)[2], 1)
    model = build_model(LinearModel(), nin, nout)

    # Objective:
    obj = EnergyDifferentialObjective(lambda=[1.0,1.0,0.0])
    generator = GenericGenerator()
    opt_state = Flux.setup(Descent(), model)
    conv = MaxIterConvergence()

    model, logs = counterfactual_training(
        obj,
        model,
        generator,
        train_set,
        opt_state;
        val_set = val_set,
        nepochs = 50,
        mutability = Symbol.(constraint),
        callback = plot_db,
        nce=50,
        convergence=conv, 
    )

    push!(models, model)
end
```

```{julia}
#| eval: false

using Plots.PlotMeasures
plts = []
titles = ["(a)", "(b)"]
_xlab = "Existing Debt"
_ylab = "Age"
for (i,model) in enumerate(models)
    coeff = get_log_reg_params(model)
    db = get_decision_boundary(coeff)
    xlab = constraints[i][1] == "both" ? "$(_xlab) (mutable)" : "$(_xlab) (immutable)"
    ylab = constraints[i][2] == "both" ? "$(_ylab) (mutable)" : "$(_ylab) (immutable)"
    plt = Plots.scatter(ce_data; xlab=xlab, ylab=ylab, bottom_margin = 5mm, left_margin = 5mm, label=["Default" "No Default"], title=titles[i])
    Plots.abline!(plt, db.slope, db.intercept; lw=5, label="Dec. Boundary")

    push!(plts, plt)
end

plt = Plots.plot(plts..., layout=(1,2), size=(600,300))
Plots.savefig(plt, "paper/figures/app_mtblty.svg")
```
